Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-10 paper.tex
terminado artigo

%
%\section{Upper and domain ontologies}\label{upper}
%There are various ontologies of interest for participatory democracy and thus for \ops. To exemplify, three sources were chosen:
%\begin{itemize}
%    \item DBpedia $\rightarrow$ ``a crowd-sourced community effort to extract structured information from Wikipedia''. Convenient for bootstrapping the OPS to out-of-domain knowledge with crowd-sourced structured information~\cite{dbpedia}.
%    \item W3C recommendations for open-government, these are the work product of specialized groups and are regarded in high account by the semantic web community~\cite{egovW3C}.
%    \item Current upper ontologies used in \ops: \foaf\ and \bfo.
%\end{itemize}
%
%This list is endless and further work should narrow the interest to essential upper ontologies. As part of the linkage to other participatory instances and even to other domains, this is a core 
%
\section{Restrictions in \vcps\ and initial \ops\ which were removed for the time being}\label{ap:restr}

These restrictions were part of \vcps\ documentation, but not implemented in \vcps\ \owl\ code:
\begin{itemize}
    \item Restriction: {\tt ops:Role ops:hasParticipationCharacteristic some ops:ParticipationCharacteristic}.
    \item Restriction: {\tt ops:Results ops:composesSolution some ops:Solution}.
    \item Restriction: {\tt Problem generatesCause some Cause}.
\end{itemize}

Present in preliminary \ops\ implementation, all restrictions were removed from current \ops\ as explained in Section~\ref{impl}
\begin{table}
  \centering
  \caption{Restrictions of the OSP (Ontology of Social Participation). All OSP restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & startsCause        & Cause\\\hline
Supporter    & supportsCause      & Cause\\\hline
Executer     & executesAction     & Action\\\hline
Solution     & solvesProblem      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & producesResults    & Results\\\hline
Result       & composesSolution   & Solution\\\hline
Cause        & dasAction          & Action\\\hline
Action       & belongsToScope     & Scope\\\hline
Cause        & hasTheme           & Theme\\\hline
Cause        & proposesSolution   & Solution\\\hline
Problem      & generatesCause     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}

Such \owl\ restrictions are valid, to the best of chances, for the final state of a participatory process (for example, in an arbitrary snapshot, a \texttt{SocialActor} may be not tied to a role). This might lead to NULL or ``not yet defined'' field supplies.
Also, in \vcps, existential restrictions were  written as ``min 1''. 
These were changed to the standard ``some'' existential restriction in preliminary \ops\ and were completely removed afterwards, as a way to avoid making \ops\ usage rigid and a more complicated.

\section{Public consultation}\label{ap:public}
Public consultations have core importance in participatory democracy and the model in figure~\ref{fig:consult} was conceived as part of \vcps\ community documentation.
An \owl\ implementation of such public consultation was meant to be developed and made public, with dedicated attention.
Considerations of this public consultation with respect to the \ops\ and formal existent instances was envisioned as well.
This public consultation model is observed here for completeness of exposition and further details goes beyond the scope of this article. A public consultations along the lines of a recent Presidential decree~\cite{decree} was formalized in \obs\ \owl\ and in \vbs\ \skos~\cite{pnud5}.

\begin{figure*}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{figs/publicConsultation__}
    \caption{A diagram representation of a general public consultation. Further context of this image is in Appendix~\ref{ap:public}.}
    \label{fig:consult}
\end{figure*}

@@

\section{Original diagram of \ops}\label{ap:dia}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.8\textwidth]{figs/diagramaOriginal}
    \caption{Original diagram of \ops. Although there were substantial OWL modifications, the main difference in the diagram is that the class \texttt{Role} was removed, as the \texttt{Role} is not able to start, execute or support anything. That is done by the Social Actor, as exposed in Figure~\ref{fig:v1}.}
    \label{fig:diaorig}
\end{figure*}

\section{Script for constructing current \ops}\label{ap:script}
Preliminary \ops\ was done with Protege software~\cite{protege}. Current \ops\ is the output of a Python script~\cite{opsScript}. Actually, this article, latex files, the \owl\ code, the \python script and auxiliary files are in a public git repository~\cite{opsRepo}, as is the common practice of the first author of this article.

@@

>>>>

%
%\section{Upper and domain ontologies}\label{upper}
%There are various ontologies of interest for participatory democracy and thus for \ops. To exemplify, three sources were chosen:
%\begin{itemize}
%    \item DBpedia $\rightarrow$ ``a crowd-sourced community effort to extract structured information from Wikipedia''. Convenient for bootstrapping the OPS to out-of-domain knowledge with crowd-sourced structured information~\cite{dbpedia}.
%    \item W3C recommendations for open-government, these are the work product of specialized groups and are regarded in high account by the semantic web community~\cite{egovW3C}.
%    \item Current upper ontologies used in \ops: \foaf\ and \bfo.
%\end{itemize}
%
%This list is endless and further work should narrow the interest to essential upper ontologies. As part of the linkage to other participatory instances and even to other domains, this is a core 
%
\section{Restrictions in VCPS and initial OPS which were removed for the time being}\label{ap:restr}

These restrictions were part of \vcps\ documentation, but not implemented in \vcps\ \owl\ code:
\begin{itemize}
    \item Restriction: {\tt ops:Role ops:hasParticipationCharacteristic some ops:ParticipationCharacteristic}.
    \item Restriction: {\tt ops:Results ops:composesSolution some ops:Solution}.
    \item Restriction: {\tt ops:Problem ops:generatesCause some ops:Cause}.
\end{itemize}

Present in preliminary \ops\ implementation, all restrictions were removed from current \ops\ as explained in Section~\ref{impl}.
\begin{table}[!h]
  \centering
  \caption{Restrictions of the \ops\ (Ontology of Social Participation). All \ops\ restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & starts        & Cause\\\hline
Supporter    & supports      & Cause\\\hline
Executer     & executes     & Action\\\hline
Solution     & solves      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & produces    & Results\\\hline
Result       & contributesTo   & Solution\\\hline
Cause        & action          & Action\\\hline
Action       & belongsTo     & Scope\\\hline
Cause        & theme           & Theme\\\hline
Cause        & proposes   & Solution\\\hline
Problem      & generates     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}

Such \owl\ restrictions are valid, to the best of chances, for the final state of a participatory process (for example, in an arbitrary snapshot, a \texttt{SocialActor} may be not tied to a role). This might lead to NULL or ``not yet defined'' field supplies.
Also, in \vcps, existential restrictions were  written as ``min 1''. 
These were changed to the standard ``some'' existential restriction in preliminary \ops\ and were completely removed afterwards, as a way to avoid making \ops\ usage rigid and more complicated.

\section{Public consultation}\label{ap:public}
Public consultations have core importance in participatory democracy and the model in Figure~\ref{fig:consult} was conceived as part of \vcps\ community documentation.
An \owl\ implementation of such public consultation was meant to be developed and made public, with dedicated attention.
Considerations of this public consultation with respect to the \ops\ and formal existent instances was envisioned as well.
This public consultation model is observed here for completeness of exposition and further details goes beyond the scope of this article. A public consultations along the lines of a recent Presidential decree~\cite{decree} was formalized in \obs\ \owl\ and in \vbs\ \skos~\cite{pnud5}.

\begin{figure*}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{figs/publicConsultation__}
    \caption{A diagram representation of a general public consultation, part of initial \vcps\ documentation. Further context of this image is in Appendix~\ref{ap:public}.}
    \label{fig:consult}
\end{figure*}

@@

\section{Original diagram of \ops}\label{ap:dia}
Although deprecated, Figure~\ref{fig:diaorig} is the basis of \vcps\ and \ops. It is displayed here for completeness of exposition and as support for discussions within this article. 
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.8\textwidth]{figs/diagramaOriginal}
    \caption{Original \vcps diagram. Although there were substantial \owl\ modifications, the main difference in the diagram is that the class \texttt{Role} was removed, as a \texttt{Role} is not able to start, execute or support anything. That is done by the Social Actor, as exposed in Figure~\ref{fig:v1}.}
    \label{fig:diaorig}
\end{figure*}

\section{Script for constructing current \ops}\label{ap:script}
Preliminary \ops\ was done with \protege\ software~\cite{protege}. Current \ops\ is the output of a Python script~\cite{opsScript}. Actually, this article, latex files, the \owl\ code, the \python script and auxiliary files are in a public git repository~\cite{opsRepo}, as is the common practice of the first author of this article.

@@

The subjects involved in this article have numerous jargons, acronyms and abbreviations. Table~\ref{tab:glos} should help readers as a concise references.
